---
title: "Midterm Project Slides"
author: "Gahyun (Grace) Kim"
date: "10/21/2019"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(wbstats)
library(tidyverse)
library(knitr)
library(magrittr)
library(esquisse)
library(rattle)
library(corrplot)
library(cluster)
library(factoextra)
library(FactoMineR)

load("/Users/home/Documents/MSSP Fall 2019/MA615/MA615_midterm/Clean_WB_data.RData")
```

## Intro

World Bank Data from R

- 1960 - 2018 records from countries worldwide
- GDP, life expectancy, urban population, etc. 

## Cleaned Data

Countries: 
```{r}
unique(data$country)
```

Variables: 
```{r}
names(data)
```

NAs were removed for most of analysis. 

## Correlation Plot

```{r}
data_scaled <- scale(data[-1])

data_corr <- cor(data_scaled)
colnames(data_corr) <- c("Date", "CO2 Emission", "GDP", "Infant Mortality", "Life Expectancy", 
                         "N2O Emission", "Nurses/Midwives per 1000", "Physicians per 1000", 
                         "Under 5 Mortality per 1000", "Urban Population", 
                         "Urban Population Growth %")
rownames(data_corr) <- c("Date", "CO2 Emission", "GDP", "Infant Mortality", "Life Expectancy", 
                         "N2O Emission", "Nurses/Midwives per 1000", "Physicians per 1000", 
                         "Under 5 Mortality per 1000", "Urban Population", 
                         "Urban Population Growth %")
corrplot(data_corr, method = "color")
```

## Scree Plot

```{r, echo = FALSE}
wssplot <- function(data, nc = 15, seed = 1234){
  wss <- nrow((data)-1)*sum(apply(data, 2, var))
  for (i in 2:nc){
    set.seed(seed)
    wss[i] <- sum(kmeans(data, centers = i)$withinss)}
  plot(1:nc, wss, type = "b", xlab = "Number of Clusters", 
       ylab = "Within groups sum of squares")}

wssplot(data_scaled, nc = 6)
```

## Cluster Analysis

```{r, echo = FALSE}
# hierarchical clustering via euclidean distance
d <- dist(data_scaled, method = "euclidean")

H.fit <- hclust(d, method = "ward")
plot(H.fit)

groups <- cutree(H.fit, k = 3)

rect.hclust(H.fit, k = 3, border = "red")
```

## Cluster Analysis

```{r, echo = FALSE}
res.dist <- get_dist(data_scaled, stand = TRUE, method = "pearson")
fviz_nbclust(data_scaled, kmeans, method = "gap_stat")
```

## Cluster Analysis

```{r, echo = FALSE}
scdata_3 <- kmeans(data_scaled, 3)
clusplot(data_scaled, scdata_3$cluster, 
         main = '2D plot of Clusters: 3 clusters', 
         color = TRUE, shade = TRUE, 
         labels = 2, lines = 0)
```

## EDA for Variables

Positively correlated variable pair 

```{r, echo = FALSE}
ggplot(data, aes(x = infant_mortality, y = under5_mortality_per_1000)) + 
  geom_point(aes(color = country)) + 
  stat_smooth(geom="line", alpha=0.3, size=0.3) + 
  labs(x = "Infant Mortality", y = "Under 5 Mortality per 1000")

```

## EDA for Variables

Negatively correlated variable pair 

```{r}
# negatively correlated variables
ggplot(data, aes(x = infant_mortality, y = life_expectancy)) + 
  geom_point(aes(color = country)) + 
  stat_smooth(geom="line", alpha=0.3, size=0.3) + 
  labs(x = "Infant Mortality", y = "Life Expectancy")
```

## EDA for Variables

Negatively correlated variable pair 

```{r}
ggplot(data, aes(x = life_expectancy, y = under5_mortality_per_1000)) + 
  geom_point(aes(color = country)) + 
  stat_smooth(geom="line", alpha=0.3, size=0.3) + 
  labs(x = "Life Expectancy", y = "Under 5 Mortality per 1000")
```

## EDA for Variables

Negatively correlated variable pair 

```{r}
ggplot(data, aes(x = infant_mortality, y = physicians_per_1000)) + 
  geom_point(aes(color = country)) + 
  stat_smooth(geom="line", alpha=0.3, size=0.3) + 
  labs(x = "Infant Mortality", y = "Physicians per 1000")
```
